---
title: "biotrace: R integration for the BioTrace GitHub Action"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{biotrace: R integration for the BioTrace GitHub Action}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`biotrace` is the R integration package for the [BioTrace](https://github.com/rotsl/biotrace) GitHub Action. BioTrace traces biological results from data and code to figures and scientific claims. It runs as a GitHub Action referenced by `uses: rotsl/biotrace@v1` in a workflow file.

This R package does not ship or run the upstream TypeScript engine. The engine runs only through GitHub. The package provides:

- R helpers to write and validate a BioTrace configuration YAML file
- R helpers to write the GitHub workflow YAML that calls the upstream Action
- An R reader for the JSON report BioTrace produces
- A command-line interface (`biotrace`) that wraps the same R API
- A conservative GitHub integration helper that can trigger an existing BioTrace workflow

```{r setup}
library(biotrace)
```

# Quick start in a fresh project

```{r, eval = FALSE}
# Initialise .github/biotrace.yml and .github/workflows/biotrace.yml
use_biotrace()
```

`use_biotrace()` writes two files and refuses to overwrite them. The workflow file references the upstream Action as `uses: rotsl/biotrace@v1`. To regenerate after editing, pass `overwrite = TRUE`:

```{r, eval = FALSE}
use_biotrace(overwrite = TRUE)
```

# Configuration

The smallest valid BioTrace configuration is:

```yaml
version: 1
```

```{r}
default_biotrace_config()
default_biotrace_config_text()
```

You can validate any BioTrace configuration file against the upstream JSON Schema (a verbatim copy of the schema published by `rotsl/biotrace` ships under `inst/extdata/schemas/`):

```{r, eval = FALSE}
validate_biotrace_config(".github/biotrace.yml")
```

A validation failure produces a structured error pointing at the offending field. Pass `return_errors = TRUE` to collect every violation instead of throwing on the first one.

# Workflows

`biotrace_workflow_text()` returns the workflow YAML as a character vector. It is deterministic and parsable by `yaml::yaml.load()`:

```{r}
cat(biotrace_workflow_text(), sep = "\n")
```

# Reports

When BioTrace runs in CI it writes a JSON report. `read_biotrace_report()` reads that file into an S3 object of class `biotrace_report`:

```{r}
report <- read_biotrace_report(
  system.file("extdata", "example-report.json", package = "biotrace")
)
print(report)
summary(report)
```

# GitHub integration

`run_biotrace_github()` is a thin wrapper around the GitHub REST API. It can list workflows, list recent runs of the BioTrace workflow, or dispatch the workflow. It reads a token from `GITHUB_TOKEN` (or `GH_TOKEN`, or any env var you name) and never writes the token to disk.

```{r, eval = FALSE}
# Preview the request that would be issued, without sending it.
run_biotrace_github("rotsl/biotrace-r",
                    action = "trigger",
                    dry_run = TRUE)
```

# Command line interface

After installation, the `biotrace` command is available on `PATH`:

```sh
biotrace --help
biotrace --version
biotrace init
biotrace config validate .github/biotrace.yml
biotrace report biotrace-report.json
biotrace github run --repo rotsl/biotrace-r --action trigger --dry-run
```

The CLI does not provide a `biotrace check` subcommand. There is no local execution of the upstream TypeScript engine in this package, so offering a `check` command would be misleading. The closest equivalent is `biotrace github run`, which talks to GitHub.
